---
title: "Anthro 490 final project"
output:
  html_document:
    df_print: paged
---
McKenna Farmer  
3/17/21  
See complete documentation on [github](https://github.com/mckfarm/anthro490)

Basic workflow steps:  
- Transform QIIME2 outputs into usable format for phyloseq  
- Subset and visualize data using phyloseq, ggplot2, and base R commands  

### Set up
```{r}
# wd and packages
setwd("~/1 School and Research/1 Northwestern/5 Classes/5 Winter 2021/Anthro 490/Project/analysis")

library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(cowplot)
library(ggcorrplot)

# functions - use this later to transform QIIME outputs to matrix
make.matrix <- function(x) {
  m<-as.matrix(x[,-1])
  rownames(m)<-x[,1]
  m
}

```


### Import QIIME2 outputs and transform
Thank you github commenters:  
- https://gist.github.com/erictleung/eda33bceceebb190eb9a44c93a077d32  
- https://github.com/joey711/phyloseq/issues/821#issuecomment-371701469
```{r}
# otu - converted table_merged.qza -> feature-table.biom -> feature-table.txt
otu <- read.table(file = "feature-table.txt", header = TRUE) 
#	tax - converted taxonomy.qza -> taxonomy.tsv
tax_midas	<- read.table(file = "taxonomy_midas.tsv", sep = '\t', header = TRUE)

# taxa cleaning
tax_midas <- separate(tax_midas, Taxon, 
                      c("Kingdom","Phylum","Class","Order", "Family", "Genus","Species"), 
                      sep= ";", remove=TRUE)

tax_midas <- sapply(tax_midas, function(x) gsub("[a-zA-Z]__","",as.character(x))) # clean up taxa names

# merge files
merged_midas <- merge(otu, tax_midas, by.x="FeatureID")

# documenting merged files
write.table(merged_midas, file = "otu_tax_midas.txt", sep = "\t", col.names = TRUE, row.names = FALSE)

```

### Make phyloseq object
```{r}
# phyloseq import
TREE <- read_tree("rooted_tree.nwk")
## had to change original metadata file from preprocessing, replaced - with . in sampleids
META <- import_qiime_sample_data("metadata.txt")
META$date <- as.Date(META$date, format = "%m/%d/%Y")
META$month <- as.factor(META$month)

# stuff we just made in this script
otu_matrix <- make.matrix(otu)
tax_midas_matrix <- make.matrix(tax_midas)
OTU <- otu_table(otu_matrix,taxa_are_rows=TRUE)
TAX_midas <- tax_table(tax_midas_matrix)

# make phylo object
physeq_midas <- phyloseq(OTU,TAX_midas,TREE,META)

# clean up workspace
rm(TREE,OTU,TAX_midas,tax_midas,tax_midas_matrix,
   otu,otu_matrix,merged_midas)
```

### Process phyloseq object before further analysis
```{r}
# removing fab's samples on fab's suggestion, sequence quality was not good
sabba_samples <- META[grep("Sabba",META$sample_id)]
sabba_samples <- sabba_samples$sample_id
midas_notfab <- subset_samples(physeq_midas, !(sample_id %in% sabba_samples))

# removing anaerobic samples for consistency
midas_notfab_aer <- subset_samples(midas_notfab, zone!="ana")

# traditional reactor - S = traditional
midas_e <- subset_samples(midas_notfab_aer, location=="S")
midas_e <- subset_taxa(midas_e, Genus!="NA") # remove NA genus

# sidestream reacotr - S2 = sidestream
midas_s2 <- subset_samples(midas_notfab_aer, location=="S2")
midas_s2 <- subset_taxa(midas_s2, Genus!="NA") # remove NA genus 

# relative abundances
rel_e <- transform_sample_counts(midas_e, function(x) x*100/sum(x))
rel_s2 <- transform_sample_counts(midas_s2, function(x) x*100/sum(x))
```

### Subsetting phyloseq object for analysis
```{r}
# get most abundant taxa
topN <- 10
most_abundant_taxa_e_aer <- sort(taxa_sums(rel_e), decreasing=TRUE)[1:topN]
rel_e_10 <- prune_taxa(names(most_abundant_taxa_e_aer), rel_e)

most_abundant_taxa_s2_aer <- sort(taxa_sums(rel_s2), decreasing=TRUE)[1:topN]
rel_s2_10 <- prune_taxa(names(most_abundant_taxa_s2_aer), rel_s2)

# document
write.csv(rel_e_10@tax_table,"top10tax_e.csv")
write.csv(rel_s2_10@tax_table,"top10tax_s2.csv")

# PAO genera - Accumulibacter, Tetrasphaera, Midas 99
rel_e_pao <- subset_taxa(rel_e,Genus==" Ca_Accumulibacter" | 
                           Genus==" Tetrasphaera" | 
                           Genus==" Halomonas")
rel_s2_pao <- subset_taxa(rel_s2,Genus==" Ca_Accumulibacter" | 
                            Genus==" Tetrasphaera" | 
                            Genus==" Halomonas")
rel_e_pao <- tax_glom(rel_e_pao,"Genus")
rel_s2_pao <- tax_glom(rel_s2_pao,"Genus")

# GAO genera - Competibacter, Defluviicoccus, Propionivibrio
rel_e_gao <- subset_taxa(rel_e,Genus==" Ca_Competibacter" | 
                           Genus==" Defluviicoccus" | 
                           Genus==" Propionivibrio" | 
                           Genus==" Ca Contendobacter" | 
                           Genus==" Micropruina")
rel_s2_gao <- subset_taxa(rel_s2,Genus==" Ca_Competibacter" | 
                            Genus==" Defluviicoccus" | 
                            Genus==" Propionivibrio" | 
                            Genus==" Ca Contendobacter" | 
                            Genus==" Micropruina")
rel_e_gao <- tax_glom(rel_e_gao,"Genus")
rel_s2_gao <- tax_glom(rel_s2_gao,"Genus")


# PAO and GAO in one phyloseq object for correlation analysis
rel_e_both <- subset_taxa(rel_e,Genus==" Ca_Accumulibacter" | 
                            Genus==" Tetrasphaera" | 
                            Genus==" Halomonas" | 
                            Genus==" Ca_Competibacter" | 
                            Genus==" Defluviicoccus" | 
                            Genus==" Propionivibrio" | 
                            Genus==" Ca Contendobacter" | 
                            Genus==" Micropruina")
rel_s2_both <- subset_taxa(rel_s2,Genus==" Ca_Accumulibacter" | 
                             Genus==" Tetrasphaera" | 
                             Genus==" Halomonas" | 
                             Genus==" Ca_Competibacter" | 
                             Genus==" Defluviicoccus" | 
                             Genus==" Propionivibrio" | 
                             Genus==" Ca Contendobacter" | 
                             Genus==" Micropruina")

# summarize to genus level
rel_e_both <- tax_glom(rel_e_both, taxrank="Genus") 
rel_s2_both <- tax_glom(rel_s2_both, taxrank="Genus")

# melt to df
rel_e_both_df <- psmelt(rel_e_both) 
rel_s2_both_df <- psmelt(rel_s2_both)

# subset columns
rel_e_both_df <- rel_e_both_df %>% select("Abundance","date","Genus") 
rel_s2_both_df <- rel_s2_both_df %>% select("Abundance","date","Genus")

# reshape to wide format for correlation function
rel_e_both_df <- spread(rel_e_both_df, key = Genus, value = Abundance) 
rel_s2_both_df <- spread(rel_s2_both_df, key = Genus, value = Abundance) 

# save correlation matrix
cor_rel_e <- cor(rel_e_both_df[-1]) 
cor_rel_s2 <- cor(rel_s2_both_df[-1])

# save significance matrix
cor_e_pmat <- cor_pmat(rel_e_both_df[-1]) 
cor_s2_pmat <- cor_pmat(rel_s2_both_df[-1])
```

### Plotting
```{r}
# top 10 genera relative abundance
abundance_e_10 <- plot_bar(rel_e_10, x="date", fill="Genus")
abundance_e_10 <- abundance_e_10 + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0)) +   
  scale_fill_viridis_d() +
  scale_y_continuous(limits = c(0,80),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")

abundance_s2_10 <- plot_bar(rel_s2_10, x="date", fill="Genus")
abundance_s2_10 <- abundance_s2_10 + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0)) +   
  scale_fill_viridis_d(option="plasma") +
  scale_y_continuous(limits = c(0,60),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")

# PAO abundance
abundance_e_pao <- plot_bar(rel_e_pao, x="date", fill="Genus") 
abundance_e_pao <- abundance_e_pao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0), legend.position="none") +   
  scale_fill_viridis_d() + 
  scale_y_continuous(limits = c(0,5),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")

abundance_s2_pao <- plot_bar(rel_s2_pao, x="date", fill="Genus")
abundance_s2_pao <- abundance_s2_pao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0), legend.position=c(.75,.8)) +
  scale_fill_viridis_d() +
  scale_y_continuous(limits = c(0,5),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("") + 
  xlab("Date")

abundance_pao_grid <- 
  plot_grid(abundance_e_pao, abundance_s2_pao,
            align="hv",nrow=1, axis="l",
            labels = c("A","B",""), label_size = 12)

# GAO abundance
abundance_e_gao <- plot_bar(rel_e_gao, x="date", fill="Genus")
abundance_e_gao <- abundance_e_gao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0), legend.position=c(.4,.8)) +
  scale_fill_viridis_d(option="plasma") +
  scale_y_continuous(limits = c(0,1),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")

abundance_s2_gao <- plot_bar(rel_s2_gao, x="date", fill="Genus")
abundance_s2_gao <- abundance_s2_gao + theme_classic() +  
  theme(axis.text.x = element_text(angle = 0), legend.position="none") +
  scale_fill_viridis_d(option="plasma") +
  scale_y_continuous(limits = c(0,1),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("") + 
  xlab("Date")

abundance_gao_grid <- plot_grid(abundance_e_gao, abundance_s2_gao,
                                align="hv",nrow=1, axis="l",
                                labels = c("A","B",""), label_size = 12)

# diversity indices
shannon <- plot_richness(midas_notfab_aer, x="hsom", measures=c("Shannon"),color="location")
shannon$layers <- shannon$layers[-1]
shannon <- shannon + theme_classic() + 
  geom_point(data=shannon$data, aes(x=hsom, y=value, color=location),alpha=0.5) +
  geom_boxplot(data=shannon$data, aes(x=hsom, y=value, color=location), alpha=0.1) + 
  theme(axis.text.x = element_text(angle = 0)) + 
  scale_color_manual(name="Reactor", labels=c("Traditional","Sidestream"),
                       values=c("#482677FF","#3CBB75FF")) +
  scale_x_discrete(limits=c("50%","25%","0%")) +
  xlab("% External Carbon") +
  ylab("")
alpha_legend <- get_legend(shannon)
shannon <- shannon + theme(legend.position="none")

invsimp <- plot_richness(midas_notfab_aer, x="hsom", measures=c("InvSimpson"),color="location")
invsimp$layers <- invsimp$layers[-1]
invsimp <- invsimp + theme_classic() + 
  geom_point(data=invsimp$data, aes(x=hsom, y=value, color=location),alpha=0.5) +
  geom_boxplot(data=invsimp$data, aes(x=hsom, y=value, color=location), alpha=0.1) + 
  theme(axis.text.x = element_text(angle = 0),
        legend.position="none") + 
  scale_color_manual(name="Reactor", labels=c("Traditional","Sidestream"),
                     values=c("#482677FF","#3CBB75FF")) +
  scale_x_discrete(limits=c("50%","25%","0%")) +
  xlab("% External Carbon") +
  ylab("") 

alpha_grid <- plot_grid(shannon, invsimp, alpha_legend, align="hv",
                        nrow=1, axis="l", rel_widths=c(1,1,0.4),
                        labels = c("A","B",""), label_size = 12)

# beta diversity via PCoA
ord <- ordinate(midas_notfab_aer, "PCoA", "bray")
pcoa <- plot_ordination(midas_notfab_aer, ord, 
                        color="hsom", shape="location", label="month")
pcoa$layers <- pcoa$layers[-1] # remove default layer because it can't be customized
pcoa <- pcoa + theme_bw() + geom_point(size=2) +
  labs(color="% External Carbon") + 
  scale_shape_discrete(name="Reactor",labels=c("Traditional","Sidestream"))+
  scale_color_viridis_d()

# correlation plots
cor_e <- ggcorrplot(cor_rel_e, p.mat=cor_e_pmat, type="lower", lab="true")
cor_s2 <- ggcorrplot(cor_rel_s2, p.mat=cor_s2_pmat, type="lower", lab="true")

```

### Plots for presentation
```{r}
# PAO abundance
abundance_e_pao <- plot_bar(rel_e_pao, x="date", fill="Genus") 
abundance_e_pao <- abundance_e_pao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0), legend.position="none") +   
  scale_fill_viridis_d() + 
  scale_y_continuous(limits = c(0,5),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")

abundance_s2_pao <- plot_bar(rel_s2_pao, x="date", fill="Genus")
abundance_s2_pao <- abundance_s2_pao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0)) +
  scale_fill_viridis_d() +
  scale_y_continuous(limits = c(0,5),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("") + 
  xlab("Date")
pao_legend <- get_legend(abundance_s2_pao)
abundance_s2_pao <- abundance_s2_pao + theme(legend.position="none") 

abundance_pao_grid <- 
  plot_grid(abundance_e_pao, abundance_s2_pao, pao_legend,
            align="hv",nrow=1, axis="l", rel_widths=c(1,1,0.5),
            labels = c("Traditional","Sidestream",""), label_size = 12)


# GAO abundance
abundance_e_gao <- plot_bar(rel_e_gao, x="date", fill="Genus")
abundance_e_gao <- abundance_e_gao + theme_classic() + 
  theme(axis.text.x = element_text(angle = 0)) +
  scale_fill_viridis_d(option="plasma") +
  scale_y_continuous(limits = c(0,1),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("Relative Abundance") + 
  xlab("Date")
gao_legend <- get_legend(abundance_e_gao)
abundance_e_gao <- abundance_e_gao + theme(legend.position="none") 

abundance_s2_gao <- plot_bar(rel_s2_gao, x="date", fill="Genus")
abundance_s2_gao <- abundance_s2_gao + theme_classic() +  
  theme(axis.text.x = element_text(angle = 0), legend.position="none") +
  scale_fill_viridis_d(option="plasma") +
  scale_y_continuous(limits = c(0,1),expand=expansion(mult = c(0,0.1))) +
  scale_x_date(date_labels="%m-%y") +
  ylab("") + 
  xlab("Date")

abundance_gao_grid <- plot_grid(abundance_e_gao, abundance_s2_gao,gao_legend,
                                align="hv",nrow=1, axis="l", rel_widths=c(1,1,0.5),
                                labels = c("Traditional","Sidestream",""), label_size = 12,
                                label_x=0.05)

# correlation plots
cor_e <- ggcorrplot(cor_rel_e, p.mat=cor_e_pmat, type="lower", lab="true",insig="blank")
cor_s2 <- ggcorrplot(cor_rel_s2, p.mat=cor_s2_pmat, type="lower", lab="true",insig="blank")
```

### Save plots
```{r}
ggsave("abundance_10_e.png",plot=abundance_e_10,width=6,height=4)
ggsave("abundance_10_s2.png",plot=abundance_s2_10,width=6,height=4)
# ggsave("abundance_pao_grid.png",plot=abundance_pao_grid,width=6,height=3.5)
# ggsave("abundance_gao_grid.png",plot=abundance_gao_grid,width=6,height=3.5)
ggsave("alpha_grid.png",plot=alpha_grid,width=6.5,height=3.5)
ggsave("pcoa_original.png",plot=pcoa,width=6.2,height=4)
ggsave("cor_e.png",plot=cor_e,width=5,height=5,dpi=320)
ggsave("cor_s2.png",plot=cor_s2,width=5,height=5,dpi=320)

ggsave("abundance_pao_grid_pres.png",plot=abundance_pao_grid,width=10,height=5,dpi=320)
ggsave("abundance_gao_grid_pres.png",plot=abundance_gao_grid,width=10,height=5,dpi=320)


```

